---
title: "Claremont CBC data analysis"
author: ""
date: ""
output: 
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

# Overview

[Claremont, CA](http://www.pomonavalleyaudubon.org/uploads/3/4/9/5/34958940/2018_2019_winter.pdf) has had a local winter [Christmas Bird Count (CBC for short)](https://www.audubon.org/conservation/history-christmas-bird-count) station since 1972. We are going to interact with the CBC data for four focal species:

* `COHA` - Cooper's Hawk
* `ACWO` - Acorn woodpecker
* `SPTO` - Spotted towhee
* `YRWA` - Yellow-rumped warbler

In the following exercises, please discuss as a group:

* What pattern, if any, do you observe for your species (or pair of species)?
    + How is their population abundance, proxied by CBC data, changing through time?
* What differences do you see between the plots for `Number` and `NumNormal`? 
* What processes do you think drive these differences?
    + What factors might be driven by the CBC itself?

Note that in the assignment [ClaremontCBC](https://rstudio.cloud/spaces/207567/project/3569413), you'll see a `ClaremontCBC.R` script that you can use to follow along with this discussion activity.

# Loading packages

First, let's load the packages that we'll need.

```{r}
###==================================================
### Loading packages & keys
###==================================================
library("dplyr")    # load package for wrangling data
library("readr")    # package for reading tabular data
library("ggplot2")  # package for plotting
```

## Importing the data

```{r}
###==================================================
### Importing the data
###==================================================
claremontCBC <- read_tsv("https://raw.githubusercontent.com/chchang47/BIOL104PO/master/data/ClaremontCBCdata.txt") # pulling in the spreadsheet from this internet location and storing it in claremontCBC

### Inspecting the data
claremontCBC %>%
  head() %>%
  View()

# The above is equivalent to:
# View( head( claremontCBC )) # delete the leading # to run
```

The Claremont CBC data has these four columns:

* `Species`: the shortened code for each species
* `Year`: the year of the CBC observation
* `Number`: the number of each species observed in that year
* `NumNormal`: `Number`/effort, where effort is the number of people participating that year

# Filtering the data

Your task is to:

* Generate a plot for one (or more) species
* Contrast the `Number` and `NumNormal` columns

First, let's consider how to filter the data to just one of the species.

```{r}
### Create your subset data
clareSub <- claremontCBC %>%
  filter(Species=="...") # select one of the 4 species and replace the ... with its short code name

### View your subset data
clareSub
```

## Generating a plot

Of the different plots (see [Week2b](https://chchang47.github.io/BIOL104PO/Week2b.html) FMI), because we have observations of distinct numeric x- and y-variables (x = time and y = species counts), a scatterplot would be appropriate.

```{r}
### Creating a scatterplot
  # Note that we will store the scatterplot in an object named "p" for plot
p <- ggplot(clareSub, aes(x=Year, y=Number)) # specifying an aesthetic where x is Year and y is the count of that species
p <- p + geom_point() # adding on points to create a scatterplot
p <- p + labs(x="...", y="...") # changing the default x and y axis labels to more informative labels (replace the ... with your text)
p # calling p which will be displayed in the plot viewer

### Slightly more advanced move to save plots using commands
  # ?ggsave # uncomment and run to load the help page
#ggsave(plot= p, filename="BirdPlot.jpeg") # replace filename with something more informative; remove leading # to run
```

### Exercise

How would you repeat the plot above but have `NumNormal` on the y-axis? Why might this be more appropriate in some cases?

## Plotting two or more species

We'll start by modifying the data filter to include two species.

First, let's consider how to filter the data to just one of the species.

```{r}
### Create your subset data
clareSub <- claremontCBC %>%
  filter(Species=="..." | Species=="...") # select two of the 4 species and replace the ... with their short code name
  # NB: the | symbol means OR - match species 1 OR species 2

### View your subset data
clareSub
```

## Generating a plot

Of the different plots (see [Week2b](https://chchang47.github.io/BIOL104PO/Week2b.html) FMI), because we have observations of distinct numeric x- and y-variables (x = time and y = species counts), a scatterplot would be appropriate.

```{r}
### Creating a scatterplot
  # Note that we will store the scatterplot in an object named "p" for plot
p <- ggplot(clareSub, aes(x=Year, y=Number, color=Species)) # specifying an aesthetic where x is Year and y is the count of that species
  # We're adding on another dimension to the visualization, which is using color to differentiate species
p <- p + geom_point() # adding on points to create a scatterplot
p <- p + labs(x="...", y="...") # changing the default x and y axis labels to more informative labels (replace the ... with your text)
p <- p + theme_classic() # choosing a simpler black and white background
p # calling p which will be displayed in the plot viewer

### Slightly more advanced move to save plots using commands
  # ?ggsave # uncomment and run to load the help page
#ggsave(plot= p, filename="BirdPlot2.jpeg") # replace filename with something more informative; remove leading # to run
```

### Exercise

How would you repeat the plot above but have `NumNormal` on the y-axis? Why might this be more appropriate in some cases?

# Answers to exercises

## Single-species version

```{r}
### Creating a scatterplot
  # Note that we will store the scatterplot in an object named "p" for plot
p <- ggplot(clareSub, aes(x=Year, y=NumNormal)) # specifying an aesthetic where x is Year and y is the count of that species
p <- p + geom_point() # adding on points to create a scatterplot
p <- p + labs(x="Year", y="Count (normalized)") # changing the default x and y axis labels to more informative labels (replace the ... with your text)
p # calling p which will be displayed in the plot viewer

### Slightly more advanced move to save plots using commands
  # ?ggsave # uncomment and run to load the help page
#ggsave(plot= p, filename="BirdPlot.jpeg") # replace filename with something more informative; remove leading # to run
```

## Multi-species version

```{r}
### Creating a scatterplot
  # Note that we will store the scatterplot in an object named "p" for plot
p <- ggplot(clareSub, aes(x=Year, y=NumNormal, color=Species)) # specifying an aesthetic where x is Year and y is the count of that species
  # We're adding on another dimension to the visualization, which is using color to differentiate species
p <- p + geom_point() # adding on points to create a scatterplot
p <- p + labs(x="Year", y="Count (normalized)") # changing the default x and y axis labels to more informative labels (replace the ... with your text)
p <- p + theme_classic() # choosing a simpler black and white background
p # calling p which will be displayed in the plot viewer

### Slightly more advanced move to save plots using commands
  # ?ggsave # uncomment and run to load the help page
#ggsave(plot= p, filename="BirdPlot2.jpeg") # replace filename with something more informative; remove leading # to run
```