---
title: "Acoustic ecology - studying environments through sound"
author: ""
date: ""
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

html_tag_audio <- function(file, type = c("wav")) {
  type <- match.arg(type)
  htmltools::tags$audio(
    controls = "",
    htmltools::tags$source(
      src = file,
      type = glue::glue("audio/{type}", type = type)
    )
  )
}
```

# Accessing the assignment

Please navigate to the assignment [Acoustic](https://posit.cloud/spaces/323650/content/5199743) in the [ConBio SP2023 workspace](https://sso.posit.cloud/pomona?redirect=https%3A%2F%2Fposit.cloud%2Fspaces%2F323650%2Fjoin%3Faccess_code%3D_klYqlJJLC2qolGKC7GmdPDI6m-xGiZnJ7wesTsQ). In the `Files` pane (typically, bottom right-hand corner), you will find an R script that you can use to follow along with this tutorial called `Acoustic.R`.

Recall that:

* You can use `Tools` --> `Global options` --> `Code` --> click `Soft-wrap R source files` to get word wrap enabled for `R` scripts.
* You need to highlight (or place your cursor on a line) and run a line (or lines) of code to execute commands.
* You can tell that the code has been executed when it is echoed (printed out) in the console.

# Overview {.tabset .tabset-fade}

In this tutorial, we'll take a look at sound data collected at the Bernard Field Station.

In this tutorial, you will learn:

* what are different ways to interact with sound data in `R`?
* how can we visualize sound data using spectrograms?
* how can we calculate indices of acoustic diversity?
* what can we do with processed sound data?

#  A brief description of sound

When we vocalize, we are making vibrations that move through the air like a wave. Microphones detect these vibrations. These vibrations are then recorded into numbers which can be processed by computers. It turns out that we can get quite a bit of information from these numbers.

Let's start by loading the packages that we will need.

```{r, message=FALSE, warning=FALSE}
### Loading packages
library(seewave)
library(tuneR)
library(dplyr)
library(readr)
```

## Waveforms

Let's pull in one of the recorded files from the BFS. You can play the recording below.

```{r,echo=FALSE}
html_tag_audio('20220313_063000.WAV')
```

Now, let's visualize this recording using an oscillogram. The output from the function `oscillo` shows how how the amplitude (or loudness) changes over the whole recording. We're going to focus on visualizing the recording from 1:01 to 1:05

```{r,echo=TRUE}
## Read in the sound data
BFS_am_1 <-tuneR::readWave('20220313_063000.WAV')
## Show the oscillogram
## which is a plot of amplitude over time
## We'll focus on the first minute of the recording
oscillo(BFS_am_1, from = 0, to = 60, fastdisp=TRUE)
```

One thing to note is that these data were recorded at a frequency of 48,000 records per second. So even though each sound file is only 5 minutes long, it has a total of 14,400,000 data points! (This is equal to 5 minutes * 60 seconds / minute * 48000 records/second.)

## Spectrogram

Another way that we can visualize sound is to use the fact that sound has the property of frequency (pitch) at particular times. We can also include amplitude (loudness) as a color. A spectrogram let's us display all three elements of an acoustic signal: its time/duration, the frequency (pitch) of the sound, and the loudness of the sound at each frequency (pitch).

We see that the frequencies in this sound recording are mostly located from 0 to 5000 Hz (or 5 kHz).

```{r,echo=TRUE,warning=FALSE,message=FALSE}
## Use spectro to display the spectogram of this data
spectro(BFS_am_1, fastdisp=TRUE)
```

# Soundscape indices

From observational data of species, we can calculate metrics such as species richness, Shannon's index, etc. Similarly, from sound data, we can calculate metrics of acoustic complexity, which theoretically should be related to the overall biodiversity of vocalizing animals.

First, let's now also read in an evening recording.

```{r,echo=FALSE}
html_tag_audio('20220313_205000.WAV')
```

```{r}
## Read in the sound data for the evening
BFS_pm_1 <-tuneR::readWave('20220313_205000.WAV')
```

Here, we will use the function `ACI` to calculate the acoustic complexity of a morning versus evening recording from the BFS.

```{r}
ACI(BFS_am_1)
ACI(BFS_pm_1)
```

# Processed data: Species richness

In addition to using numeric metrics such as `ACI` to calculate the acoustic diversity of audio recordings, we can use new computer vision models to identify (bird) species in the recordings. This therefore gives us data that we tend to see in conservation biology: **species richness**. We have and will process/ed the raw recording data to:

* Calculate ACI for each recording;
* Identify bird vocalizations in each recording;

We have used Cornell Lab of Ornithology's [BirdNET](https://birdnet.cornell.edu/) neural network model, which is trained on data such as identified acoustic recordings from [xeno-canto](https://xeno-canto.org/). This model extracts distinct bird vocalizations from the spectrographic representation of each sound file. Those bird vocalizations are then pattern matched against known bird vocalizations to generate the most likely species (of the 3000 species supported by the model) with some probability (a.k.a. `confidence`).

Let's take a look at some example data.

```{r}
# Read in a spreadsheet from a URL using read_tsv from the readr package
birdnetDF <- readr::read_tsv("https://github.com/EA30POM/site/raw/main/data/birdnet10-22.tsv")

# Inspect the data
birdnetDF %>% head() # look at the first 6 rows
```

Below, I'll walk you through some possible analyses with the BirdNET identifications, so that this can be grist for the mill for thinking about a hypothesis that your group can take on. Let's start by removing low confidence recordings (e.g. confidence $ < 0.5$).

```{r}
# Filter out low confidence identifications
birdnetDF <- birdnetDF %>%
  dplyr::filter(Confidence >= 0.5) # only retain the rows (observations) where Confidence >= 0.5
```

Now, let's say that we were interested in evaluating the number of species identified during the day or night. How would we do that?

```{r}
# Analyzing number of species during the day or night
birdnetDF %>%
  group_by(DayNight) %>% # group by the Day or Night categories
  summarize(NumSpecies = length(unique(CommonName))) # number of distinct species per category
```

Below is a summary of what the relevant columns in the class data are or will be:

*`birdnet` Data*:

* `SpeciesCode`: A [six letter short-hand code for species](https://www.birdpop.org/pages/birdSpeciesCodes.php), determined by groups such as the American Ornithological Society
* `CommonName`: The common name for different species identified in each 3 second window of the recording
* `Confidence`: The confidence of the identification for each species in each 3 second window
* `unit`: Which unit recorded the data
* `date`: What was the date of the recording?
* `time`: What was the time of the recording?
* `DayNight`: Was the recording made during the daytime or nighttime?

## Your task

Your job is to think through as groups a question that you would be interested in answering using processed acoustic data from the BFS that will span from October-November 2022 and our current data in March-April 2023. Here are some suggestions about external datasets that you can look at:

* [BirdCast](https://dashboard.birdcast.info/region/US-CA-037)
* [NOAA NCDC Claremont Station](https://www.ncei.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USR0000CCLA/detail)
* [CA Dept Water Resources, historic temperature data](https://cdec.water.ca.gov/dynamicapp/QueryDaily?s=CMO&end=2022-07-06)
* [CA Data Exchange-CMO station](https://cdec.water.ca.gov/cgi-progs/plotReal?staid=CMO)
* [CIMIS data](https://cimis.water.ca.gov/)
* [Google Earth map of sampling stations](https://earth.google.com/earth/d/1-8gneVsVX1zM89AHJQhloNBqVLUZUYnH?usp=share_link)

# `R` script

Below is the `R` script in its entirety.

```{r, eval=F}
###===============================================================
### Section 3
###===============================================================

### Loading packages
library(seewave)
library(tuneR)
library(dplyr)
library(readr)

###===============================================================
### Section 3.1 - 3.2
###===============================================================

## Read in the sound data
BFS_am_1 <-tuneR::readWave('BFS_sound_2022/20220313_063000.WAV')

## Show the oscillogram
## which is a plot of amplitude over time
## We'll focus on the first minute of the recording
oscillo(BFS_am_1, from = 0, to = 60, fastdisp=TRUE)

## Use spectro to display the spectogram of this data
spectro(BFS_am_1, fastdisp=TRUE)

###===============================================================
### Section 4
###===============================================================

## Read in the sound data for the evening
BFS_pm_1 <-tuneR::readWave('BFS_sound_2022/20220313_205000.WAV')

### Morning sound recording acoustic complexity index
ACI(BFS_am_1)

### Evening sound recording acoustic complexity index
ACI(BFS_pm_1)

###===============================================================
### Section 5
###===============================================================

# Read in a spreadsheet from a URL using read_tsv from the readr package
birdnetDF <- readr::read_tsv("https://github.com/EA30POM/site/raw/main/data/birdnet10-22.tsv")

# Inspect the data
birdnetDF %>% head() # look at the first 6 rows

# Filter out low confidence identifications
birdnetDF <- birdnetDF %>%
  dplyr::filter(Confidence >= 0.5) # only retain the rows (observations) where Confidence >= 0.5

# Analyzing number of species during the day or night
birdnetDF %>%
  group_by(DayNight) %>% # group by the Day or Night categories
  summarize(NumSpecies = length(unique(CommonName))) # number of distinct species per category
```