---
title: "The species-area relationship"
author: ""
date: ""
output: 
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE, results = FALSE, fig.show = FALSE)
```

# Species-area relationship

Observations of species richness (counts of species) across islands (or habitat fragments--sometimes called patches) showed a **consistent** relationship where **as islands became larger in area, they tended to support more species**: that is, *larger habitat areas have higher species richness*. This is a pattern that has been consistent across different taxonomic groups (e.g. vascular plants, birds, insects, fish) and regions (e.g. California, Southeast Asia, the Florida keys, etc.).

The species-area relationship holds that we can *predict* the number of species that would be found on a habitat of a given *area*. Specifically, if $S$ denotes the number of species (species richness), then $S = c A^z$ where $A$ is habitat area, and $c$ and $z$ are constants.

```{r SEasia_birds}
### Loading packages
library(dplyr)
library(tibble)
library(ggplot2)
POMblue <- "#0057B7" # Pomona blue hex code

### Code to create a data table storing the Southeast Asian bird data
seAsiaBirds <- tibble::tibble(areaMi2=c(312000,290000,144000,70000,48000,25000,4500,8870,18000,4600),
            birds=c(540,420,368,220,337,232,111,143,137,108))

### Display first few rows of the Southeast Asia bird dataset
seAsiaBirds

### Create a plot of bird richness versus island area
p <- ggplot(seAsiaBirds,aes(x=areaMi2, y=birds))
p <- p + geom_point()
p <- p + labs(x="Area (square miles)",y="Bird species richness")
p <- p + theme_classic()
p
```

Oof! What we can see in the previous plot are the issues I raised in class on Tuesday (lecture 9). Issue 1): The x-axis values and y-axis values are quite wide-ranging (x-axis range: [`r paste(signif(range(seAsiaBirds$areaMi2),5),collapse=", ")`]; y-axis range: [`r paste(range(seAsiaBirds$birds),collapse=", ")`]). Issue 2): The relationship looks like a non-linear relationship, and a power law specifically. Is there some way that we can make this relationship of `species richness ~ habitat area` more linear?

## Calculating $d$ and $z$

### Log-log transforming the data

I show you how to calculate $d$ (where $d = \log_{10} c$ from the log-log linear species-area relationship) and $z$ in `R` below. `lm` is an `R` function to perform an [ordinary least squares linear regression](https://www.r-bloggers.com/2017/07/ordinary-least-squares-ols-linear-regression-in-r/). Don't worry too much about the details for `lm`; we'll look at `lm` in more depth in a later week's tutorial.

First, we confirm that the log-log transformation makes the Southeast Asian bird-habitat area relationship look linear. Second, we use `lm` to calculate $d$ and $z$.

```{r SEasia_fit_c_z}
### Below, we first apply log10 to the columns of our data on SE Asian birds
seAsiaBirds <- seAsiaBirds %>%
  mutate(log10Area = log10(areaMi2), log10Birds = log10(birds))

  # Display seAsiaBirds in the console
seAsiaBirds

  # Confirm that the log-log transformation linearized our data
p <- ggplot(seAsiaBirds,aes(x = log10Area, y = log10Birds))
p <- p + geom_point()
p <- p + labs(x="Area (log-transformed)",y="Birds (log-transformed)")
p <- p + theme_classic()
p
```

### Calculating $d$ and $z$

Below, we will calculate the coefficients for the species-area relationship.

```{r}
### Use lm to calculate the values of d and z and store the model in the object speciesAreaModel
speciesAreaModel <- lm(log10Birds ~ log10Area, data=seAsiaBirds)
  ## See the values estimates for d and z
print("The cofficient names (Intercept) corresponds to d and log10Area to z")
coef(speciesAreaModel)
  ## Store the estimates in the objects d and z
d <- coef(speciesAreaModel)[1] # first item in this two-item vector is the intercept of the straight line fit to our log-log transformed Southeast Asian birds dataset
z <- coef(speciesAreaModel)[2] # second item in the two-item vector is the slope
```

Let's try an example! Given these values for $z=0.35$ and $d=0.76$ (where $c=10^d = 10^{0.76} = 5.7$), let's estimate how many birds we would expect for the smallest island, Palawan: $S = 10^{0.76} 4500~\text{square miles}^{0.35}$, which yields around `r floor(10^0.76 * 4500^0.35)` species. In our dataset, Palawan in fact had 111 species - so we can see that this predicted richness of 109 species is not too far off from the actual richness!

### Exercise:

* How many species would you expect for the island of Celebes (70000 $mi^2$)?

# Calculating the number of species remaining after habitat loss

For the discussion activity, assume we're looking at habitat loss in Palawan. Assume it is 100% forested. What would happen if we lose 10% of the forest habitat in Palawan?

```{r species_after_loss}
### We can take the value of z that we calculated in the Southeast Asia
### bird dataset and use that to calculate 1) what proportion of species
### would remain after habitat loss and 2) what number of species would be
### lost after habitat loss.
SR1 = 111 # initially we have 111 species in this example (Palawan)
R = 0.9 # We lose 10% of habitat in this example: (1-R)*100% = 10%
species_proportion_surviving <- R^z # what proportion of species will persist?
species_proportion_surviving # print out answer

### How many species will remain?
remaining_species_richness <- SR1 * R^z
remaining_species_richness # print out answer
```

## Exercise

* How many (forest) species will be lost in Palawan with this 10% reduction in forest habitat area?
* Repeat the calculations above for a 20% reduction in habitat for Palawan.

# Solutions

*  How many species would you expect for the island of Celebes (70000 $mi^2$)?

```
habitat_area <- 70000 # Celebes' land area in square miles
sr_exp <- 10^d*habitat_area^z # calculating the expected species richness given the prior values
sr_exp # we see the number of expected species
```

* How many (forest) species will be lost in Palawan with this 10% reduction in forest habitat area?

```
species_lost <- SR1 * (1-R^z)
species_lost
```

* Repeat the calculations above for a 20% reduction in habitat for Palawan.

```
SR1 = 111 # initially we have 111 species in this example (Palawan)
R = 0.8 # We lose 20% of habitat in this example: (1-R)*100% = 20%
species_proportion_surviving <- R^z # what proportion of species will persist?
species_proportion_surviving # print out answer

### How many species will remain?
remaining_species_richness <- SR1 * R^z
remaining_species_richness # print out answer

### How many species will be lost?
species_lost <- SR1 * (1-R^z)
species_lost
```

# Additional context

Below, I provide a lengthier explanation both about 1) how the log-log transformation takes the non-linear species-area relationship and makes it linear, and 2) how and why we have the equation presented in class for calculating species losses expected from habitat loss.

## More background: the log-log species-area relationship

For the relationship between species and habitat area, $S = c A^z$, how do we go about estimating $c$ and $z$? This expression is a power law, so we can make it linear instead by taking a logarithm (henceforth, log) of both sides, such that: $\log_{10}(S) = \log_{10}(c A^z)$ 

[New England Complex Systems Institute](https://necsi.edu/power-law#:~:text=An%20example%20is%20the%20area,of%20a%20power%20law%20relationship.) has a great explainer for *power law* relationships and a familiar example: the area of a square, which is always a power of its length. We can turn this power law relationship between habitat area and species richness to a linear one, by taking the logs of both variables (area and species richness).

[MathBench at Univ. of Maryland](https://mathbench.umd.edu/modules/misc_scaling/page11.htm) has a great, intuitive explanation for why log-log transformations "linearize" power laws (that is, turn the relationship from a power law into a straight line).

Can we simplify this algebra above? Yeah! Note that $\log_{10}(c A^z)$, by the properties of logs (see below if you'd like refresher on log properties), can be re-expressed as $\log_{10}(c) + \log_{10}(A^z)$, which can be further simplified to $\log_{10}(c) + z \log_{10}(A)$. From here on, I'm going to be lazy and drop the $()$ and just write the above expression as $\log_{10} c + z \log_{10} A$.

Now we can take $\log_{10} c + z \log_{10} A$, and substitute it in to the right-hand side of $\log_{10}(S) = \log_{10}(c A^z)$, which yields $\log_{10} S = \log_{10} c + z \log_{10} A$.

Hm ... this equation $\log_{10} S = \log_{10} c + z \log_{10} A$ looks awfully like the equation for a straight line, if we let $y$ denote $\log_{10} S$ and $x$ denote $\log_{10} A$. Recall that the equation for a straight line is $y = d + m x$, where $d$ is the y-intercept (the value of $y$ when $x=0$) and $m$ is "rise over run", or the slope, or simply how much $y$ changes with every unit change in $x$. ([A nice explainer on slopes from the Monterey Institute.](http://www.montereyinstitute.org/courses/Algebra1/U04L1T1_RESOURCE/topicText.html#:~:text=In%20math%2C%20slope%20is%20the,is%20called%20the%20slope%20formula.))

By substituting in the values from the linear species-area equation into the equation for a straight line, we realize that: $y = \log_{10} S; d + mx = \log_{10} c + z \log_{10} A$. What does that mean? By pattern matching, we see that $\log_{10} c$ is playing the same role as $d$ (y-intercept) and $z$ is playing the same role as $m$ (slope or rise-over-run).

Thus, given that the relationship between $\log_{10} S$ and $\log_{10} A$ is a straight line, we expect that we should be able to 1) use $\log_{10}$ to transform our actual species richness and habitat area data, then 2) fit a straight line to those $\log_{10} - \log_{10}$ transformed data.

### Revisiting the Timor expected species richness calculation

Now that we've estimated $d$ and $z$ above for the Southeast Asian bird dataset, let's use those values to calculate the expected species richness on the island of Timor, which is 18,000 square miles. Given that $\log_{10} S = \log_{10} c + z \log_{10} A$ and that $d = \log_{10} c$, we can directly plug in our $d$ and $z$ values into: $\log_{10} S = d + z \log_{10} A$. That's annoying though, because we get a value for $S$ that is log-10 transformed. That is, in this case $S$ is described in powers of 10, rather than the actual number of species.

Alternatively, we can get rid of the logs by raising both sides to the power of 10, yielding: $10^{(\log_{10} S)} = 10^{(d + z \log_{10} A)} \rightarrow 10^{(\log_{10} S)} = 10^{(d \log_{10}(A^z))} \rightarrow S = 10^d A^z \implies S = c A^z$. Recall that $d = \log_{10} c \implies 10^d = c$, so this ends up working out as we convert $d$ back to $c$ by raising 10 to the power of $d$.

Phew! Let's go ahead and calculate the expected richness for Timor (18000 miles$^2) = 10^d A^z$, which is about `r round(10^d*18000^z, 2)` species. The actual number of bird species on Timor is `filter(seAsiaBirds,areaMi2==18000) %>% select(birds)`, which is `r select(filter(seAsiaBirds,areaMi2==18000),birds)` species. The numbers look pretty similar! However, it does appear that Timor has somewhat fewer species than expected (137 birds when given the other Southeast Asian island-bird data points, we'd expect 176.76 species).

### Quick refresher on logarithms

Khan Academy has a great refresher series on [calculating logarithms](https://youtu.be/Z5myJ8dg_rM) and [logarithm properties](https://youtu.be/PupNgv49_WY). 

You can also confirm the different rules of logs by using `R` as a calculator! Recall that in [Week 1](https://chchang47.github.io/BIOL104PO/Week1b.html), we learned about how `R` can also be used as a calculator in addition to statistical computing. Below, I show how you can confirm the properties of logs using `R`.

Additionally, this is a friendly reminder that if you're ever confused about a function, you can pull up a help page about it using `?functionname` (e.g. `?mutate` or `?geom_point`). One of the things I find most helpful on the help page is the **examples** that are typically provided at the bottom of the help page.

```{r log10_explorations}
### Recall the log-log species-area relationship:
### log S = log(c A^z) (Equation 1)
### we can confirm that the previous equation is 
### mathematically equivalent to the following one:
### log S = log c + log(A^z) (Eqn. 2)
### which is in turn equal to:
### log S = log c + z log A (Eqn. 3)
### For convenience, we'll call c and z something different:
### c1 and z1
### That way, we don't overwrite the previously calculated values
### for c and z from the Southeast Asia bird dataset

### Let c1 = 5, z1 = 0.28, A = 1000
### We'll calculate both expressions
### and confirm that they are identical
c1 <- 5
z1 <- 0.28
A <- 1000
eqn1 <- log10(c1*A^z1)
eqn1 # value of the first equation
eqn2 <- log10(c1) + log10(A^z1)
eqn2 # value of second equation
eqn3 <- log10(c1) + z1*log10(A)
eqn3 # value of third equation
```

## More background on how to estimate species losses from habitat loss using the species-area relationship

One of the reasons why the species-area relationship is so central to conservation is that it gives us a way to estimate how habitat loss can reduce the number of species a location (a.k.a. site, patch, fragment) can support.

However, we immediately encounter a problem. If we want to predict how species richness will change at one site, and we don't have a local dataset to estimate $c$ and $z$, what are we supposed to do? Past ecological research indicates that $z$ tends to fall in a fairly narrow range between [0.15, 0.39] (Preston 1962; MacArthur and Wilson 1967). However, values for $c$ vary widely, depending on the taxonomic group (e.g. birds have a very different y-intercept for their straight-line log-log species-area relationship than insects or microbes) and region.

In a situation where we don't have values of $c$ and $z$ for our system, while we can at least lean on these past values for $z$, we're going to want some way to get rid of $c$. 

Let's start with $A_0$ denoting the **original amount (area) of habitat** and $A_1$ denoting the **amount of habitat left after some is destroyed**; that is, $A_1 < A_0$. Similarly, $S_0$ denotes the initial number of species and $S_1$ the number of species after habitat conversion has brought the habitat area down from $A_0$ to $A_1$.

From the species-area (power law) relationship, we know that:
$S_1 = c A_1^z$
$S_0 = c A_0^z$

Maybe we can get rid of the pesky $c$ term by dividing the two equations like so...
$\frac{S_1}{S_0} = \frac{c A_1^z}{c A_0^z}$

We can cancel the two $c$ terms in the numerator and denominator of the right-hand side fraction, and then the equation simplifies to $\frac{S_1}{S_0} = \frac{A_1^z}{A_0^z}$

We can simplify things further by noting that $A_1$ is always going to be some fraction of $A_0$, as $A_0$ represents the original extent (area) of habitat, and $A_1$ the amount of that habitat that is still standing after destruction happens. That is, $A_1 = R A_0$ where $R$ stands for habitat $R$emaining after loss.

In that case, when we plug in this expression $A_1 = R~A_0$ to $\frac{S_1}{S_0} = \frac{A_1^z}{A_0^z}$ we arrive at:
$\frac{S_1}{S_0} = \frac{(R~A_0)^z}{A_0^z}$

Based on the [product rule for exponents](https://www.mesacc.edu/~scotz47781/mat120/notes/exponents/review/review.html#:~:text=Product%20Rule%3A%20am%20%E2%88%99,base%20and%20add%20the%20powers.&text=%2C%20this%20says%20that%20to%20divide,base%20and%20subtract%20the%20powers), $\frac{S_1}{S_0} = \frac{(RA_0)^z}{A_0^z}$ is equivalent to:
$\frac{S_1}{S_0} = \frac{R^z A_0^z}{A_0^z}$

We can cancel out the two $A_0^z$ terms and finally land at:
$\frac{S_1}{S_0} = R^z$
This equation gives us a way to calculate **what proportion of species remain after some fraction of habitat is lost**.

We can multiply the denominator on the left hand side over, seeing that $S_1 = S_0 R^z$. We can then also calculate the **number of species lost**, $S_0 - S_1$, denoted as $\Delta S$ as:
$\Delta S = S_0 - S_1 \rightarrow S_0 - S_0 R^z \implies \Delta S = (1-R^z) S_0$.